# Generalized linear models

This chapter gives a hands-on introduction to generalized linear models with a
focus on logistic regression for binary response data and poisson regression for
count data.

## Slides

```{=html}
<object data="slides/02_glm/02_glm.pdf" type="application/pdf" width="100%" height="500px">
  <p>
    Unable to display PDF file. <a href="slides/02_glm/02_glm.pdf">Download</a> instead.
  </p>
</object>
```

## Exercises

### Exercise 1 {-}

* In a psychophysical experiment two LEDs are presented to a subject: a standard
  with 40 cd/m$^2$ and a comparison with varying intensities

* The subject is supposed to say which stimulus is brighter; each comparison is
  presented 40 times

  |               |    |    |    |    |    |    |    |
  |---------------|---:|---:|---:|---:|---:|---:|---:|
  | x (cd/m$^2$)  | 37 | 38 | 39 | 40 | 41 | 42 | 43 |
  | y (positiv)   |  2 |  3 | 10 | 25 | 34 | 36 | 39 |


* Estimate parameters $c$ and $a$ of the logistic psychometric function
  \[
    p_{pos} = \frac{1}{1 +
      \exp(-\frac{\displaystyle x - c}{\displaystyle a})}
  \]
  using `glm()´ with $logit(p_{pos}) = \beta_0 + \beta_1x$ where $a = 1/\beta_1$
  and $c = -\beta_0/\beta_1$.

* Calculate the intensity $x$ for which $p_{pos} = 0.5$ (Point of Subjective
  Equality, PSE)


* Create a plot for the probability to give a positive answer depending on the
  intensity of the comparison

* Use `predict()` to obtain the predicted values and add the logistic
  psychometric function to the plot

* Use `abline()` to add parameter $c$ to the plot

* Use a likelihood ratio test to assess how well the model fits the data

* Is there reason to assume that there is any overdispersion?

```{r ex02.1, results = "hide", class.source = "fold-hide", fig.show = "hide"}
dat <- data.frame(x = 37:43,
                  y = c(2, 3, 10, 25, 34, 36, 39),
                  n = 40)

glm1 <- glm(cbind(y, n - y) ~ x, family = binomial, data = dat)

a <- 1 / coef(glm1)[2]
c <- -coef(glm1)[1] / coef(glm1)[2]

newx <- seq(37, 43, .1)
pre <- predict(glm1, newdata = data.frame(x = newx), type = "response")

# Plot predictions with PSE
plot(y/n ~ x, data = dat, pch = 16, ylab = "Probability to say brighter")
lines(pre ~ newx, data = dat)
abline(v = c, h = .5, lty = 3)
text(39, .8, paste("PSE =", round(c, 2)))

# Goodness-of-fit test
glms <- glm(cbind(y, n - y) ~ factor(x), family = binomial, data = dat)
anova(glm1, glms, test = "Chisq")

# Overdispersion
summary(glm(cbind(y, n - y) ~ x, family = quasibinomial, data = dat))
```

### Exercise 2 {-}

* Fit a regression model to the `Affairs` data set from the `AER` package in R

* The variable `affairs` is the number of extramarital affairs in the past year
  and is our response variable

* Include the variables `gender`, `age`, `yearsmarried`, `children`,
  `religiousness`, `education` and `rating` as predictors

* `religiousness` ranges from 1 (anti) to 5 (very) and `rating` is a self rating
  of the marriage, ranging from 1 (very unhappy) to 5 (very happy)

* Assess the Goodness-of-fit using the deviance

* Assess overdispersion and decide if a model with an extra dispersion parameter
  might be indicated

* Compare the confidence intervals for the estimated parameters for both models

```{r ex02.2, results = "hide", class.source = "fold-hide", fig.show = "hide"}
# Load data set
data(Affairs, package = "AER")
Affairs$rating <- factor(Affairs$rating, levels = 1:5)

## Fit poisson model
pois1 <- glm(affairs ~ gender + age + yearsmarried + children + education +
             rating, family = poisson, data = Affairs)
summary(pois1)

# Goodness-of-fit
1 - pchisq(pois1$deviance, df = pois1$df.residual)

## Model with dispersion parameter
pois2 <- glm(affairs ~ gender + age + yearsmarried + children + education +
             rating, family = quasipoisson, data = Affairs)
summary(pois2)

## Compare CIs
confint(pois1)
confint(pois2)
```

#### Visualize model predictions {-}

```{r ex02.3, class.source = "fold-hide"}
## Plot predictions for "average" females
newdat <- expand.grid(yearsmarried = seq(0, 15, 1), rating = factor(1:5))
newdat$gender        <- "female"
newdat$age           <- mean(Affairs$age)
newdat$children      <- "yes"
newdat$religiousness <- mean(Affairs$religiousness)
newdat$education     <- mean(Affairs$education)
newdat$pre           <- predict(pois2, newdata = newdat, type = "response")

colors <- c("#78004B", "#3CB4DC", "#91C86E", "#FF6900", "#434F4F")
plot(pre ~ yearsmarried, newdat, type = "n", main = "'Average' Females")

for (i in 1:5) {
  lines(pre ~ yearsmarried, newdat[newdat$rating == i, ], col = colors[i])
}

legend("topleft", c("very unhappy", "somewhat unhappy", "average",
                    "happier than average", "very happy"),
       col = colors, lty = 1, bty = "n")
```

