# Pre/post measurements

In this chapter, we start with the simplest form of longitudinal data: pre/post
measurements.

## Slides

```{=html}
<object data="slides/03_prepost/03_prepost.pdf" type="application/pdf" width="100%" height="500px">
  <p>
    Unable to display PDF file. <a href="slides/03_prepost/03_prepost.pdf">Download</a> instead.
  </p>
</object>
```

## Exercises

### Exercise 1 {-}

@Lowry2000[Chap. 17] describes the following example

* Based on values $Y$ of a finals exam, $m = 3$ learning methods introducing
  basic programming skills will be compared ($i = 1, 2, 3$)

* In each of the chosen schools, 12 students attending grade 5 have been
  randomly selected ($j = 1, \dots, 12$) for a six week course

* Before the course, familiarity with computers was assessed in a pre test as
  covariate $X$

```{r ex03.1, echo = FALSE}
dat <- read.table("data/ancova-data.txt", header = TRUE)
dat$method <- factor(dat$method)
```

Data for method 1:
```{r ex03.2, echo = FALSE}
knitr::kable(t(dat[dat$method == 1, c("pre", "post")]), col.names = NULL)
```

Data for method 2:
```{r ex03.3, echo = FALSE}
knitr::kable(t(dat[dat$method == 2, c("pre", "post")]), col.names = NULL)
```

Data for method 3:
```{r ex03.4, echo = FALSE}
knitr::kable(t(dat[dat$method == 3, c("pre", "post")]), col.names = NULL)
```

```{r ex03.5, echo = FALSE, fig.width = 3.375, fig.height = 3.375, fig.align = "center"}
par(mai = c(.8, .8, .1, .1), mgp = c(2.4, 1, 0))
datm <- aggregate(cbind(pre, post) ~ method, data = dat, FUN = mean)
dotchart(as.matrix(datm[, 2:3]), xlab = "Score", ylab = "Method")
```

* Fit the following models to data

  \begin{align}
    y_{ij} &= \beta_0 + \beta_1 \cdot x_{ij} + \varepsilon_{ij}\\
    y_{ij} &= \beta_0 + \beta_1 \cdot x_{ij} + \beta_2 \cdot z_{i} + \varepsilon_{ij}\\
    y_{ij} &= \beta_0 + \beta_1 \cdot x_{ij} + \beta_2 \cdot z_{i} +
              \beta_3 \cdot x_{ij} \cdot z_i + \varepsilon_{ij} 
  \end{align}

* Plot the data with the predictions for each model

* Compare the models with a likelihood ratio test

* What are the adjusted means for the ANCOVA model?

```{r ex03.6, results = "hide", class.source = "fold-hide", fig.show = "hide"}
## Fit models
lm1 <- lm(post ~ pre, data = dat)
lm2 <- lm(post ~ pre + method, data = dat)
lm3 <- lm(post ~ pre * method, data = dat)

## LRT
anova(lm1, lm2, lm3)

colors <- c("#78004B", "#FF6900", "#3CB4DC")

## Plot model predictions
par(mfrow = c(1, 3))

plot(post ~ pre, data = dat, col = colors[dat$method], pch = 16)
abline(lm1)
legend("topleft", paste("Method", 1:3), col = colors, pch = 16, bty = "n")

plot(post ~ pre, data = dat, col = colors[dat$method], pch = 16)
abline(coef(lm2)[1], coef(lm2)[2], col = colors[1])
abline(coef(lm2)[1] + coef(lm2)[3], coef(lm2)[2], col = colors[2])
abline(coef(lm2)[1] + coef(lm2)[4], coef(lm2)[2], col = colors[3])
legend("topleft", paste("Method", 1:3), col = colors, pch = 16, bty = "n")

plot(post ~ pre, data = dat, col = colors[dat$method], pch = 16)
abline(coef(lm3)[1], coef(lm3)[2], col = colors[1])
abline(coef(lm3)[1] + coef(lm3)[3], coef(lm3)[2] + coef(lm3)[5], col = colors[2])
abline(coef(lm3)[1] + coef(lm3)[4], coef(lm3)[2] + coef(lm3)[6], col = colors[3])
legend("topleft", paste("Method", 1:3), col = colors, pch = 16, bty = "n")

## Means for pre and post measurements
datagg <- aggregate(cbind(pre, post) ~ method, data = dat, FUN = mean)

## Adjusted means
datagg$post_adj <- predict(lm2,
                           newdata = data.frame(method = factor(1:3),
                                                pre = mean(dat$pre)))

# Adjusted "time effects"
datagg$diff <- datagg$post - datagg$pre
datagg$diff_adj <- datagg$post_adj - mean(datagg$pre)
```

### Exercise 2 {-}

* Simulate a data set for 75 subjects

* Create a factor `condition` indicating a control group and two
  treatment groups with 25 subjects each

* Simulate pre test data with $N \sim (5, 1)$

* Simulate a post test score assuming a slope of 0.7 between pre and post score
  and that the first treatment group improves by 0.5 points compared to the
  control group and the second treatment group by 1 point

* What other assumption needs to be made?

```{r, results = "hide", class.source = "fold-hide"}
set.seed(1042)      # set seed for reproducibility

n <- 75
condition <- factor(rep(c("contr", "treat1", "treat2"), each = n / 3))
eff_cond <- c(1, 0.5, 1)

pre  <- rnorm(n, mean = 5, sd = 1)
post <- 0.7 * pre + model.matrix( ~ condition) %*% eff_cond + rnorm(n)

dat <- data.frame(id        = factor(1:n),
                  pre       = pre,
                  post      = post,
                  condition = condition
       )
rm(pre, post, condition)

lattice::xyplot(post ~ pre, dat, groups = condition,
                type = c("g", "p", "r"), auto.key = TRUE)

# Observed averaged time effects, pre and post scores
aggregate(cbind(post - pre, pre, post) ~ condition, dat, mean)
```

* Fit three different models to the data
  1. A Change Score model
  2. An ANCOVA model
  3. A mixed-effects model with predictors `condition` and `time`

* Compare the results of the three models

* What is the adjusted time effect for each group?

```{r, results = "hide", class.source = "fold-hide"}
library(lme4)

## Change Score Model

#m1 <- lm(post - pre ~ condition, dat)
m1 <- lm(post ~ condition + offset(pre), dat)
summary(m1)

## ANCOVA Model

m2 <- lm(post ~ condition + pre, dat)
summary(m2)

# Adjusted means
predict(m2, newdata = data.frame(condition = c("contr", "treat1", "treat2"),
                                 pre = mean(dat$pre)))

## Mixed Model (equivalent to change score model)

dat_long <- reshape(dat, direction = "long",
                    varying = list(c("pre", "post")),
                    v.names = "resp",
                    idvar = "id",
                    timevar = "time",
                    times = c(0, 1))

m3 <- lmer(resp ~ condition*time + (1 | id), dat_long)
summary(m3)

## Compare parameters of the three models
cbind(coef(m1), coef(m2)[1:3], fixef(m3)[4:6])

datagg <- aggregate(cbind(pre, post) ~ condition, data = dat, FUN = mean)

## Adjusted means
datagg$post_adj <- predict(m2,
                           newdata = data.frame(condition = factor(c("contr", "treat1", "treat2")),
                                                pre = mean(dat$pre)))

# Adjusted "time effects"
datagg$diff <- datagg$post - datagg$pre
datagg$diff_adj <- datagg$post_adj - mean(datagg$pre)
```

### Exercise 3 {-}

@Hedeker2006 report a smoking prevention study with students. The study used a
$2 \times 2$ factorial design with factor "prevention curriculum" (with vs.
without) and "TV prevention" (with vs. without). Students were randomly assigned
to one of the four groups. Students' knowledge about tabacco induced health
risks before and after the prevention have been measured.

* Read the data set `television.txt` into R. Assign meaningful variable names.
  (You will find the necessary information at the beginning of the data file!)

* Calculate means and standard deviations for baseline, follow-up and change
  scores separately for the four groups. How many students are assigned to each
  group?

* Plot the follow-up means. Put intervention curriculum on the $y$ axis and two
  seperate lines for each tv condition. Which effects do you expect based on
  this plot?

* Conduct an analysis for the follow-up scores (ANOVA), the change scores, the
  adjusted follow-up scores (ANCOVA), and the adjusted change score. Interpret
  the parameters for these models. How does the knowledge of the students change
  depending on the two factors?

* Assess how well the assumptions for the ANCOVA model hold using visual
  methods.

```{r, results = "hide", class.source = "fold-hide", fig.show = "hide"}
dat <- read.table("data/television.txt", skip = 43)
names(dat) <- c("school", "class", "curri", "tv", "pre", "post")

aggregate(cbind(pre, post, post - pre) ~ curri + tv, data = dat, FUN = mean)
aggregate(cbind(pre, post) ~ curri + tv, data = dat, FUN = sd)
aggregate(pre ~ curri + tv, data = dat, FUN = length)

interaction.plot(dat$curri, dat$tv, dat$post, type = "b", pch = c(1,16),
  xlab = "Curriculum", ylab = "Mean THKS score", trace.label = "TV")

summary(lm1 <- lm(post ~ curri * tv, data = dat))              # Follow-Up
summary(lm2 <- lm(post - pre ~ curri * tv, data = dat))        # Change-Score
summary(lm3 <- lm(post ~ pre + curri * tv, data = dat))        # ANCOVA
summary(lm4 <- lm(post - pre ~ pre + curri * tv, data = dat))  # Change-Score ANCOVA

plot(lm3)
```

