# Repeated measurements

## Slides

```{=html}
<object data="slides/05_repeated_measures/05_repeated_measures.pdf" type="application/pdf" width="100%" height="500px">
  <p>
    Unable to display PDF file. <a href="slides/05_repeated_measures/05_repeated_measures.pdf">Download</a> instead.
  </p>
</object>
```

## Exercises

* Expand the linear mixed-effects model two more than one factor

* Add diagnosis ("endogenous" vs. "non-endogenous") as additional between factor

* Test if this factor interacts with `week` using a likelihood-ratio test

* Use parametric bootstrapping to get a sampling distribution for your LRT
  statistic

* Plot the sampling distribution and add the empirical confidence interval

* Plot the predictions of the model

  \[
    y_{ij} = \beta_0 + \beta_1 time + \beta_2 diag + \upsilon_{0i} + \upsilon_{1i} time +
    \varepsilon_{ij}
  \]
  with
  \begin{align*}
    \begin{pmatrix} \upsilon_{0i}\\ \upsilon_{1i} \end{pmatrix} &\sim
      N \left(\begin{pmatrix} 0\\ 0 \end{pmatrix}, \, \boldsymbol{\Sigma}_\upsilon =
        \begin{pmatrix}
          \sigma^2_{\upsilon_0} & \sigma_{\upsilon_0 \upsilon_1} \\
          \sigma_{\upsilon_0 \upsilon_1} & \sigma^2_{\upsilon_1} \\
        \end{pmatrix} \right)
      \text{ i.i.d.} \\
    \boldsymbol{\varepsilon}_i &\sim N(\mathbf{0}, \, \sigma^2 \mathbf{I}_{n_i})
      \text{ i.i.d.}
  \end{align*}

```{r ex05.1, results = "hide", class.source = "fold-hide"}
dat      <- read.table("data/reisby.dat", header = TRUE)
dat$id   <- factor(dat$id)
dat$diag <- factor(dat$diag, levels = c("nonen", "endog"))
dat      <- na.omit(dat)     # drop missing values

# Add diagnosis
lme1 <- lmer(hamd ~ week + diag + (week | id), dat, REML = FALSE)
lme2 <- lmer(hamd ~ week * diag + (week | id), dat, REML = FALSE)
anova(lme1, lme2)
```

```{r ex05.2, cache = TRUE, results = "hide", class.source = "fold-hide", fig.show = "hide"}
# Parametric bootstrapping
nsim <- 1000
LR <- rep(NA, nsim)
sim2 <- simulate(lme2, nsim)

for (i in seq_len(nsim)) {
  fit1 <- lmer(sim2[, i] ~ week + diag + (week | id), dat, REML = FALSE)
  fit2 <- lmer(sim2[, i] ~ week * diag + (week | id), dat, REML = FALSE)
  LR[i] <- 2 * (summary(fit2)$logLik - summary(fit1)$logLik)
}

hist(LR, breaks = 50, col = "grey", border = "white", freq = FALSE, main = "")
curve(dchisq(x, df = 1), add = TRUE)
abline(v = 2 * (summary(lme2)$logLik - summary(lme1)$logLik), lty = 2)
# Empirical p-value
mean(LR > 2 * (summary(lme2)$logLik - summary(lme1)$logLik))
```

```{r ex05.3, results = "hide", class.source = "fold-hide", fig.show = "hide"}
# Predictions with diagnosis as fixed effect
datm <- aggregate(hamd ~ week + diag, data = dat, FUN = mean)
datm$pred1 <- predict(lme1, newdata = datm, re.form = ~0)

colors <- c("#3CB4DC", "#91C86E")

plot(hamd ~ week, data = datm, col = colors[datm$diag],
     ylim = c(0, 28), xlab = "Week", ylab = "HDRS score")
lines(pred1 ~ week, data = subset(datm, datm$diag == "nonen"), col = colors[1])
lines(pred1 ~ week, data = subset(datm, datm$diag == "endog"), col = colors[2])
legend("bottomleft", legend = c("Endogenous", "Non endogenous"), col = colors,
       pch = 21, lty = 1, pt.bg = "white", bty = "n")

```

