# Crossed random effects

## Slides


```{=html}
<object data="slides/07_crossed_random/07_crossed_random.pdf" type="application/pdf" width="100%" height="500px">
  <p>
    Unable to display PDF file. <a href="slides/07_crossed_random/07_crossed_random.pdf">Download</a> instead.
  </p>
</object>
```

## Exercises

### Exercise 1 {-}

* Change the data simulation by Baayen2008 for $N = 30$ subjects
  instead of only 3

* You can use the following script and adjust it accordingly

* You can choose if you want to use model matrices or create the vectors
  "manually"


```{r ex07.1}
library(lattice)
library(lme4)

#--------------- (1) Create data frame ---------------
datsim <- expand.grid(subject = factor(c("s1" , "s2" , "s3" )),
                      item    = factor(c("w1" , "w2" , "w3" )),
                      soa     = factor(c("long" , "short" )))
datsim <- datsim |> sort_by(~ subject)

#--------------- (2) Define parameters ---------------
beta0 <- 522.11
beta1 <- -18.89

sw  <- 21.1
sy0 <- 23.89
sy1 <- 9
ry  <- -1
se  <- 9.9

#--------------- (3) Create vectors and simulate data ---------------
# Fixed effects
b0 <- rep(beta0, 18)
b1 <- rep(rep(c(0, beta1), each = 3), 3)

# Draw random effects
w <- rep(rnorm(3, mean = 0, sd = sw), 6)
e <- rnorm(18, mean = 0, sd = se)

# Bivariate normal distribution
sig <- matrix(c(sy0^2, ry * sy0 * sy1, ry * sy0 * sy1, sy1^2), 2, 2)
y01 <- mvtnorm::rmvnorm(3, mean = c(0, 0), sigma = sig)
y0 <- rep(y01[,1], each = 6)
y1 <- rep(c(0, y01[1,2],
            0, y01[2,2],
            0, y01[3,2]), each = 3)

datsim$rt <- b0 + b1 + w + y0 + y1 + e

#--------------- (4) Simulate data using model matrices ---------------
X <- model.matrix( ~ soa, datsim)
Z <- model.matrix( ~ 0 + item + subject + subject:soa, datsim,
                  contrasts.arg = list(subject = contrasts(datsim$subject,
                                                        contrasts = FALSE)))

# Fixed effects
beta <- c(beta0, beta1)
# Random effects
u <- c(w = unique(w),
       y0 = y01[,1],
       y1 = y01[,2])

datsim$rt2 <- X %*% beta + Z %*% u + e

#--------------- (5) Visualize simulated data ---------------
xyplot(rt ~ soa | subject, datsim, group = item, type = "b", layout = c(3, 1))
```


```{r ex07.2, results = "hide", class.source = "fold-hide", fig.show = "hide"}
n <- 30

datsim <- expand.grid(subject = factor(paste0("s", 1:n)),
                      item    = factor(c("w1" , "w2" , "w3" )),
                      soa     = factor(c("long" , "short" )))
datsim <- datsim |> sort_by(~ subject)

beta0 <- 522.11
beta1 <- -18.89

sw  <- 21.1
sy0 <- 23.89
sy1 <- 9
ry  <- -1
se  <- 9.9

w <- rnorm(3, mean = 0, sd = sw)
e <- rnorm(n * 6, mean = 0, sd = se)

# Bivariate normal distribution
sig <- matrix(c(sy0^2, ry * sy0 * sy1, ry * sy0 * sy1, sy1^2), 2, 2)
y01 <- mvtnorm::rmvnorm(n, mean = c(0, 0), sigma = sig)

beta <- c(beta0, beta1)
# Random effects
u <- c(w = w,
       y0 = y01[,1],
       y1 = y01[,2])

X <- model.matrix( ~ soa, datsim)
Z <- model.matrix( ~ 0 + item + subject + subject:soa, datsim,
                  contrasts.arg = list(subject = contrasts(datsim$subject,
                                                        contrasts = FALSE)))

datsim$rt <- X %*% beta + Z %*% u + e

xyplot(rt ~ soa | subject, datsim, group = item, type = "b", layout=c(5, 6))
```

### Exercise 2 {-}

* Fit the following models to the healing data by @Aungle2023

```{r ex07.3}
m1 <- lmer(Healing ~ Condition + (1 | Subject) + (1 | ResponseId), dat)
m2 <- lmer(Healing ~ Condition + (Condition | Subject) + (1 | ResponseId), dat)
m3 <- lmer(Healing ~ Condition + (1 | Subject) +
           (0 + dummy(Condition, "28") | Subject) +
           (0 + dummy(Condition, "56") | Subject) +
           (1 | ResponseId), dat)
```

* Profile the models with `profile(<model>)`

* Use the functions `xyplot()`, `densityplot()`, `splom()` from the lattice
  package to take a closer look at the estimated random parameters

* Compare the three models with likelihood ratio tests

* What is the best model in your opinion?


```{r ex07.4, results = "hide", class.source = "fold-hide", fig.show = "hide"}
pm1 <- profile(m1)
pm2 <- profile(m2)
pm3 <- profile(m3)

ci1 <- confint(pm1)
ci2 <- confint(pm2)
ci3 <- confint(pm3)

xy1 <- ci1[c("(Intercept)", "Condition28", "Condition56"), ]
xy2 <- ci2[c("(Intercept)", "Condition28", "Condition56"), ]
xy3 <- ci3[c("(Intercept)", "Condition28", "Condition56"), ]

dp <- data.frame(par = c("(Intercept)", "Condition28", "Condition56"),
                 model = rep(c("m1", "m2", "m3"), each = 3),
                 ci = c(xy1[, 1], xy2[, 1], xy3[, 1], xy1[, 2], xy2[, 2], xy3[, 2])
)

dotplot(par ~ ci, dp[dp$model != "m3",], groups = model, auto.key = T, xlab = "")

dotplot(model ~ ci | par, dp, groups = model, xlab = "", layout = c(1, 3), pch = 16)
```
