# Contrast coding

## Slides

```{=html}
<object data="slides/09_contrast_coding/09_contrast_coding.pdf" type="application/pdf" width="100%" height="500px">
  <p>
    Unable to display PDF file. <a href="slides/09_contrast_coding/09_contrast_coding.pdf">Download</a> instead.
  </p>
</object>
```

## Exercises

### Exercise 1 {-}

* Simulate data based on the example by @Brehm2022 for a 2x2 between-subjects
  design

```{r ex09.1, results = "hide", class.source = "fold-hide", fig.show = "hide"}
library(lme4)
set.seed(1012)

# Average time to eat in min
SpoonSoup  <- 5
ForkSoup   <- 10
SpoonSalad <- 10
ForkSalad  <- 5

# Standard deviation for all groups
Groupsd <- 2

# Number of subjects (ps) and ingredients (ii) - crossed random design
ps <- 20
ii <- 10

ds <- data.frame(
  Utensils    = factor(rep(c("Spoon", "Fork"), each = ps * ii, times = 2)),
  Foods       = factor(rep(c("Soup", "Salad"), each = ps * ii * 2)),
  Participant = factor(rep(paste0("p", 1:ps), times = ii * 4)),
  Item        = factor(rep(paste0("i", 1:ii), each = ps, times = 4)))

ds$RT <- c(rnorm(ps * ii, mean = SpoonSoup,  sd = Groupsd),
           rnorm(ps * ii, mean = ForkSoup,   sd = Groupsd),
           rnorm(ps * ii, mean = SpoonSalad, sd = Groupsd),
           rnorm(ps * ii, mean = ForkSalad,  sd = Groupsd))

psre <- rnorm(ps, mean = 0, sd = Groupsd / 5)
iire <- rnorm(ii, mean = 0, sd = Groupsd / 5)

ds$RT <- ds$RT + psre[ds$Participant] + iire[ds$Item]
```

* Fit the following model to your simulated data
  $$
    RT = \beta_0 + \beta_1 UtensilsSpoon + \beta_2 FoodsSoup + \beta_3
    UtensilsSpoon \times FoodsSoup + \upsilon_0 + \eta_0 + \varepsilon
  $$
  with $\upsilon_0 \sim N(0, \sigma_{\upsilon}^2)$, $\eta_0 \sim N(0,
  \sigma_{\eta}^2)$, $\varepsilon \sim N(0, \sigma_{\varepsilon}^2)$, all
  i.i.d.

  1. with dummy coding
  2. with effects coding

```{r ex09.2, results = "hide", class.source = "fold-hide", fig.show = "hide"}
# Treatment coding
m1 <- lmer(RT ~ Utensils * Foods + (1 | Participant) + (1 | Item), data = ds)
summary(m1)

# Effects coding
contrasts(ds$Utensils) <- contr.sum(levels(ds$Utensils))
contrasts(ds$Foods)    <- contr.sum(levels(ds$Foods))
m2 <- lmer(RT ~ Utensils * Foods + (1 | Participant) + (1 | Item), data = ds)
summary(m2)
```

* Interpret the fixed parameters and compare the variance components for the
  random effects for the two models

* What is the interpretation of the intercept for the two models? Why does it
  differ?

### Exercise 2 {-}

* Fit linear models with all contrasts we looked at

  1. Treatment contrasts
  2. Sum contrasts
  3. Helmert contrasts
  4. Sequential difference contrasts
  5. Custom contrasts

* Interpret the fixed parameters

* Predict data with each model and compare the results

* Look again at the intercepts for all models and compare

