# Generalized linear mixed-effects models

## Slides

```{=html}
<object data="slides/11_glmm/11_glmm.pdf" type="application/pdf" width="100%" height="500px">
  <p>
    Unable to display PDF file. <a href="slides/11_glmm/11_glmm.pdf">Download</a> instead.
  </p>
</object>
```

## Exercises

### Exercise 1 {-}

* Load the example data from https://github.com/eveskew/glmm_tutorial

```{r ex11.1}
# Download the data from https://github.com/eveskew/glmm_tutorial

dat <- read.csv("https://raw.githubusercontent.com/eveskew/glmm_tutorial/refs/heads/master/glmm_tutorial_data.csv")
```

* The data set contains data on 11 studies investigating how many animals got
  tested positive for a certain disease (sorry, I could not find out which one)

* Additional variables are `Habitat` indicating if the animals were tested in a
  forest (`Forest`) or in a agricultural context (`Ag`) and `Season` with levels
  `Spring` and `Fall`

* Create a plot with prevalence on the y-axis, habitat on the x-axis, two panels
  for season and one line for each study

* Fit GLMMs predicting the probability that an animal is tested positive using
  `Season` and `Habitat` as predictors

* What would you choose as random effect here?

* Which model would you choose?

* Interpret your results

```{r ex11.2, results = "hide", class.source = "fold-hide", fig.show = "hide"}
library(lme4)

dat$Habitat <- factor(dat$Habitat, levels = c("Forest", "Ag"))
dat$Season  <- factor(dat$Season)
dat$Study   <- factor(dat$Study)

dat$Prev <- dat$Positive / dat$Num_Tested

# Plot data
lattice::xyplot(Prev ~ Habitat | Season, data = dat, type = "b", groups = Study)

# Fit GLMMs
gm1 <- glmer(cbind(Positive, Num_Tested - Positive) ~ 1 + (1 | Study),
             data = dat, family = binomial)
gm2 <- glmer(cbind(Positive, Num_Tested - Positive) ~ Habitat + (1 | Study),
             data = dat, family = binomial)
gm3 <- glmer(cbind(Positive, Num_Tested - Positive) ~ Habitat + Season +
             (1 | Study),
             data = dat, family = binomial)
gm4 <- glmer(cbind(Positive, Num_Tested - Positive) ~ Habitat * Season +
             (1 | Study),
             data = dat, family = binomial)

anova(gm1, gm2, gm3, gm4)

# Odds ratio for Habitat
exp(fixef(gm2))[2]
```


