# Power simulation for LMMs

```{r}
library(lattice)
library(lme4)
```

# Reanalysis
## Application context: Depression and type of diagnosis

- @ReisbyGram77 studied the effect of Imipramin on 66 inpatients
  treated for depression
- Depression was measured with the Hamilton depression rating scale
- Patients were classified into endogenous and non-endogenous depressed
- Depression was measured weekly for 6 time points

Data: [reisby.txt](data/reisby.txt)

```{r}
dat      <- read.table("data/reisby.txt", header = TRUE)
dat$id   <- factor(dat$id)
dat$diag <- factor(dat$diag, levels = c("nonen", "endog"))
dat      <- na.omit(dat)     # drop missing values
head(dat, n = 13)

xyplot(hamd ~ week | id, data = dat, type=c("g", "r", "p"),
  pch = 16, layout = c(11, 6), ylab = "HDRS score", xlab = "Time (week)")
```

## Random-intercept model

$$
\begin{aligned}
Y_{ij} &= \beta_0 + \beta_1 \, \mathtt{week}_{ij}
        + \upsilon_{0i}
        + \varepsilon_{ij} \\
\upsilon_{0i} &\sim N(0, \sigma^2_{\upsilon_0}) \text{ i.i.d.} \\
\mathbf{\varepsilon}_i &\sim N(0, \, \sigma^2)  \text{ i.i.d.} \\
i &= 1, \ldots, I, \quad j = 1, \ldots n_i
\end{aligned}
$$

```{r}
m1 <- lmer(hamd ~ week + (1 | id), data = dat, REML = FALSE)
summary(m1)
```

## Random-slope model

$$
\begin{aligned}
Y_{ij} &= \beta_0 + \beta_1 \, \mathtt{week}_{ij}
        + \upsilon_{0i} + \upsilon_{1i}\, \mathtt{week}_{ij}
        + \varepsilon_{ij} \\
\begin{pmatrix} \upsilon_{0i}\\ \upsilon_{1i} \end{pmatrix} &\sim
   N \left(\begin{pmatrix} 0\\ 0 \end{pmatrix}, \,
           \mathbf{\Sigma}_\upsilon =
\begin{pmatrix}
          \sigma^2_{\upsilon_0} & \sigma_{\upsilon_0 \upsilon_1} \\
          \sigma_{\upsilon_0 \upsilon_1} & \sigma^2_{\upsilon_1} \\
\end{pmatrix} \right)
      \text{ i.i.d.} \\
\mathbf{\varepsilon}_i &\sim N(\mathbf{0}, \, \sigma^2 \mathbf{I}_{n_i})
      \text{ i.i.d.} \\
i &= 1, \ldots, I, \quad j = 1, \ldots n_i
\end{aligned}
$$

```{r}
m2 <- lmer(hamd ~ week + (week | id), data = dat, REML = FALSE)
summary(m2)
```

## Partial pooling

```{r fig.height = 10, fig.width = 6.5, fig.aling = "center"}
indiv <- unlist(
  sapply(unique(dat$id),
         function(i) predict(lm(hamd ~ week, dat[dat$id == i, ])))
)

xyplot(hamd + predict(m2, re.form = ~ 0) + predict(m2) + indiv ~ week | id,
  data = dat, type = c("p", "l", "l", "l"), pch = 16, grid = TRUE,
  distribute.type = TRUE, layout = c(11, 6), ylab = "HDRS score",
  xlab = "Time (week)",
  # customize colors
  col = c("#434F4F", "#3CB4DC", "#FF6900", "#78004B"),
  # add legend
  key = list(space = "top", columns = 3,
             text = list(c("Population", "Mixed model", "Within-subject")),
             lines = list(col = c("#3CB4DC", "#FF6900", "#78004B")))
  )
```

## By-group random-slope model

```{r}
m3 <- lmer(hamd ~ week + diag + (week | id), data = dat, REML = FALSE)
m4 <- lmer(hamd ~ week * diag + (week | id), data = dat, REML = FALSE)
anova(m3, m4)
```

## Means and predicted HDRS score by group

```{r}
dat2 <- aggregate(hamd ~ week + diag, dat, mean)
dat2$m4 <- predict(m4, newdata = dat2, re.form = ~ 0)

plot(m4 ~ week, dat2[dat2$diag == "endog", ], type = "l",
     ylim=c(0, 28), xlab="Week", ylab = "HDRS score")
lines(m4 ~ week, dat2[dat2$diag == "nonen", ], lty = 2)
points(hamd ~ week, dat2[dat2$diag == "endog", ], pch = 16)
points(hamd ~ week, dat2[dat2$diag == "nonen", ], pch = 21, bg = "white")
legend("topright", c("Endogenous", "Non endogenous"),
       lty = 1:2, pch = c(16, 21), pt.bg = "white", bty = "n")
```

# Power simulation

## Setup

```{r}
## Study design and sample sizes
n_week <- 6
n_subj <- 80
n <- n_week * n_subj
dat <- data.frame(
     id = factor(rep(seq_len(n_subj), each = n_week)),
   week = rep(0:(n_week - 1), times = n_subj),
  treat = factor(rep(0:1, each = n/2), labels = c("ctr", "trt"))
)

## Fixed effects and variance components
beta <- c("(Intercept)" = 23, week = -0.5, treat = 0, "week:treat" = -1)
se <- 3.5                # residual sd
r <- -0.3
t(chol(VarCorr(m4)$id))[lower.tri(diag(2), diag = TRUE)] / sigma(m3)
# su <- c(3.5, 1.5)
# Su <- r * su %o% su
# diag(Su) <- su^2        # covariance matrix of random effects

su1 <- 3.5
su2 <- 1.5
Su <- matrix(c(su1^2, r * su1 * su2, r * su1 * su2, su2^2), nrow = 2, ncol = 2)
```

## Power

### Simulate data with bivariate normal distribution

```{r cache = TRUE, warning = FALSE}
pval <- replicate(200, {
  
  # Data generation
  means <- model.matrix( ~ week * treat, dat) %*% beta
  ranu <- MASS::mvrnorm(n_subj, mu = c(0, 0), Sigma = Su)
  e <- rnorm(n_subj * n_week, mean = 0, sd = se)
  
  y <- means + ranu[dat$id, 1] + ranu[dat$id, 2] * dat$week + e
  
  # Fitting model to test H0
  m0 <- lmer(y ~ week + treat + (1 + week | id), data = dat, REML = FALSE)
  m1 <- lmer(y ~ week * treat + (1 + week | id), data = dat, REML = FALSE)
  anova(m0, m1)["m1", "Pr(>Chisq)"]
  }
)

mean(pval < 0.05)
```

### Simulate data with `lme4::simulate()`


```{r cache = TRUE, warning = FALSE}
## Gory details
fixef(m4)
getME(m4, "theta")
t(chol(VarCorr(m4)$id))[lower.tri(diag(2), diag = TRUE)] / sigma(m4)

# Cholesky decomposition
Lt <- chol(Su) / s
pars <- list(theta = t(Lt)[lower.tri(Lt, TRUE)],
             beta = beta, sigma = s)
names(pars$theta) <- c("id.(Intercept)", "id.week.(Intercept)", "id.week")

pval <- replicate(200, {
  y <- simulate(~ week * treat + (week | id),
                newparams = pars, newdata = dat)$sim_1
  m1 <- lmer(y ~ week + treat + (week | id), data = dat, REML = FALSE)
  m2 <- lmer(y ~ week * treat + (week | id), data = dat, REML = FALSE)
  anova(m1, m2)$"Pr(>Chisq)"[2]
})
mean(pval < 0.05)
```

## Parameter recovery

```{r}
par <- replicate(200, {
  
  means <- model.matrix( ~ week * treat, dat) %*% beta
  ranu <- MASS::mvrnorm(n_subj, mu = c(0, 0), Sigma = Su)
  e <- rnorm(n_subj * n_week, mean = 0, sd = se)
  
  y <- means + ranu[dat$id, 1] + ranu[dat$id, 2] * dat$week + e
  
  m1 <- lmer(y ~ week * treat + (1 + week | id), data = dat, REML = FALSE)
  list(fixef = fixef(m1), theta = getME(m1, "theta"), sigma = sigma(m1))
  }, simplify = FALSE
)

rowMeans(sapply(par, function(x) x$fixef))
rowMeans(sapply(par, function(x) x$theta))
mean(sapply(par, function(x) x$sigma))

beta
Lt <- chol(Su)
t(Lt)[lower.tri(Lt, diag = TRUE)] / se
se
```

### References

